### 2012-10-29 latex/paper.tex
Initial paper state

\documentclass[conference]{IEEEtran}
\usepackage{graphicx}
\usepackage{times}

\begin{document}

\title{Six ways to understand monads\\
\small{Experiences from the functional programming course at TU Delft}}

\author{\IEEEauthorblockN{Georgios Gousios}
\IEEEauthorblockA{
Delft University of Technology\\
Delft, The Netherlands\\
Email: G.Gousios@tudelft.nl} 
\and
\IEEEauthorblockN{Erik Meijer}
\IEEEauthorblockA{Microsoft \\
Rendmond, WA, USA\\
Email: H.J.M.Meijer@tudelft.nl} 
}   
\maketitle

\begin{abstract}

  In the last few years, the popularity of functional programming as a way
  of solving computational problems has increased significantly. While most
  computer science curricula do include a course on functional programming,
  in many cases it is disconnected from practical applications, which is
  precisely where functional programming shines. To fill in this gap, we 
  designed a functional programming course that demanded from students to learn
  by experience with real world applications. In this work, we present our
  experiences with conducting this course.
\end{abstract}

\section{Introduction}

Due to a variety of reasons, including the advent of cloud computing, the rising
rate of information production and the necessity to reach the market fast,
currently, large corporations and start-ups alike are investigating
alternative programming and information storage models. As a result, during the
last few years, the practical software engineering field is witnessing a noticeable shift towards functional programming. Scripting languages,
notably Javascript and Ruby, pioneered the introduction of functional
concepts, such as closures and lambda functions, to mainstream programming. A
new wave of programming languages, developed to overcome the expressiveness and
complexity limitations exhibited in mainstream languages, have promoted
functional constructs, such as type safe pattern matching, higher order
functions and single assignment variables, to first class citizens (Scala). New,
purely functional, languages have emerged to fill in the remaining gaps (F\#,
Clojure), often introducing significant advancements in their field of
specialisation (such as Erlang in distributed fault-tolerant systems). Finally,
large scale information processing systems such as Map-Reduce and domain
specific languages such as LINQ have integrated functional concepts to ease the
expression of computations.

Broadly speaking, functional programming is a style of programming in which the
primary method of computation is the application of functions to arguments.
Among other features, functional languages offer a compact notation for writing
programs, powerful abstraction methods for structuring them, and a simple
mathematical basis that supports reasoning. Many of the advanced techniques in
modern functional languages, such as monads and catamorphisms, are closely based
on principles from category theory such as functors, initial algebras, monads
and Kleisli categories.

While functional programming has been taught for long in computer science
departments~\cite{Joost93}, curricula tend to emphasize functional programming theory rather
than practical applications. Special programming languages are used to teach
functional programming specific concepts, while little connection is made to how
those concepts can be transfered to non-functional languages.  The course at TU
Delft attempted to teach functional programming theory in the classroom and then
expect students to apply those concepts in the (non-purely functional) language
of their choice. The course had a very strong teaching by example focus:
students were expected to participate in both in-classroom exercises, homework
assignments and implement a real world system as a final project.

\section{The Course}

The aim of the course is to teach the principles of pure functional
programming, and the corresponding Category theoretical principles, using the
Haskell programming language. More specifically, the educational purposes
of the course were:

\begin{itemize}

  \item To introduce students to basic functional programming concepts, such
    as higher order functions, monads and type inference.

  \item To introduce students to solving data processing problems in a 
    functional way.

  \item To explain the application of functional concepts in non-purely
    functional environments.

\end{itemize}

The course consisted of a series of lectures

\subsection{Student profile}

The course was elective at the MSc curriculum. Students from all departments in
the Electrical Engineering, Mathematics and Computer science faculty were
allowed to participate. Most students had received formal introduction to
imperative and object oriented programming in their bachelors curriculum, while
through participation to other courses they had limited exposure to functional
programming (i.e. Map/Reduce in the web information systems course). Two of the
students were majoring in computer engineering, which meant that their
programming experience was somewhat limited. Several students also had work
experience as programmers as part of their industrial placement or through
participation to start up ventures.

For organizational reasons, the participation to the course was limited to 15
students.

\subsection{Lectures}

\subsection{Student projects}

At the end of the lecture period, the students were given a selection of
projects to work on. The projects included:

\begin{itemize}

  \item Real time graph visualisation on steaming data. The
    particular example that students worked on was visualizing community
    structures for Github projects.

  \item Multi-source real time data processing. Students worked on a 
    programming language popularity index

  \item Implementation of Haskell constructs in Javascript. Students implemented
    the Haskell prelude (basic functions for list manipulation).

  \item Implementation of Map/Reduce algorithms in Haskell. Students installed
    and configured cloud Haskell and implemented simple Map/Reduce based
    algorithms in a distributed setting.

  \item Machine learning algorithms. Students implemented Na\"ive Bayes 
    classifiers and K-means clusters.

\end{itemize}

\section{Experiences}

\subsection{Teaching Monads}

\subsection{The teapot experiment}

One of the highlights of the taught period of the course was the teapot
experiment. While not a formal experiment in the scientific sense of the word,
we used it to draw the student's attention to the following facts:

\begin{itemize}

  \item Most modern programming languages can express functional programming
    constructs.

  \item Functional programming works best in data transformation scenarios

\end{itemize}

The exercise consisted of rendering the Utah Teapot~\cite{Torre06} using any
graphics primitive of the student's choice using only right triangles.  In its
core, the exercise required students to decompose arbitrary triangles, which
comprised the input Utah Teapot model, to a series right triangles.  The
students had to come up with the decomposition method (using some form of
analytic geometry), a decomposition restriction criterion to stop the
decomposition when triangles could not be rendered on screen and a method to
recursively apply the above mentioned transformations on the input data. As
always the students could decide the implementation language of their choice.

The students' response to the exercise was overwhelming. For four days,
the students worked hard to

\subsection{Teaching Map/Reduce}

\subsection{Student Projects}

\section{Conclusion}

\section*{Acknowledgements}

We would like to thank\ldots

\bibliography{paper}
\bibliographystyle{ieeetr}
\end{document}

### 2012-10-29 bin/pybaux.py

@@ -1,0 +1,57 @@
+# Original code by Adriaan de Groot. 
+#
+# Disributed under the BSD license
+#
+# Functions to parse TeX/LaTeX aux files
+
+import os, re, string
+
+# regular expression to match in the .aux file
+citation_re = re.compile ('\\citation\{([^\}]+)\}')
+include_re  = re.compile ('^\\\@input\{([^\}]+)\}')
+
+# report errors by throwing this exception
+class AuxError(Exception):
+    def __init__(self, value):
+        self.value = value
+    def __str__(self):
+        return repr(self.value)
+
+
+# Open an .aux file and retrun it's file handle
+def aux_open(file):
+    auxname = os.path.splitext (file) [0] + '.aux'
+    try:
+        auxf = open (auxname, 'r')
+    except IOError, err:
+        raise AuxError, '%s: %s' % (auxname, err)
+    return auxf
+
+# Readin and expand .aux file, returning a list of all lines in the auxfile and
+# all .aux files it references
+def aux_lines(file):
+    auxf = aux_open(file)
+    lines = [] 
+    # parse the whole file
+    for line in auxf.readlines():
+        line = string.strip (line)
+        lines.append(line)
+        # we have to enter an additional .aux file
+        match = include_re.search (line)
+        if match:
+            lines = lines + aux_lines (match.group (1))
+    auxf.close ()
+    return lines
+
+# Extract citation keys from the (expanded) .aux file and return as list
+def aux_citations(file):
+    citations = []
+    for line in aux_lines(file):
+        # we match a new citation
+        match = citation_re.search (line)
+        if match:
+            references = match.group(1)
+            for k in references.split(","):
+                citations.append (k.strip())
+    return citations
+

### 2012-10-29 bin/pybbib.py

@@ -1,0 +1,217 @@
+# Original code by Adriaan de Groot. 
+#
+# Disributed under the BSD license
+#
+# Functions to parse/write BiBTeX files
+
+import string, os, sys
+
+# report errors by throwing this exception
+class BibError(Exception):
+    def __init__(self, value):
+        self.value = value
+    def __str__(self):
+        return repr(self.value)
+
+
+# open a bib file
+def bib_open(file):
+    bibname = os.path.splitext (file) [0] + '.bib'
+    try:
+        bibf = open (bibname, 'r')
+    except IOError, err:
+        raise BibError, "File does not exist"
+    return bibf
+
+# Parsing/scanning bib file and keeping track of current char, line number
+# etc.
+
+# contains a single bibfile as a single string
+buffer = ""
+# pointer to the current character to be read
+cur = 0
+# line number corresponding to the current character to be read
+lineno = 0
+# last bibtex key successfully parsed
+lastsuccessfulkey = ""
+
+
+# open a bibfile and initialise the scan buffer
+def initch(file):
+    global buffer, cur, lineno
+    f = bib_open(file)
+    buffer = f.read()
+    cur = 0
+    lineno = 1
+
+# get the current character on the input, None of EOF    
+def curch():
+    global buffer, cur
+    if cur < len(buffer):
+        return buffer[cur]
+    else:
+        return None
+
+# advance to the next character on the input    
+def nextch():
+    global cur, lineno
+    if curch() == "\n":
+        lineno = lineno + 1 
+    cur = cur + 1
+    
+# skip a comment: all characters upto the next end of line
+def skip_comment():
+    while curch() <> '\n':
+        nextch()
+
+# skip whitespace; this includes comments (but not EOF).
+# curch() returns the first nonwhitespace char, or None if at EOF
+def skip_whitespace():
+    while (curch() <> None) and ((curch() == "%") or (curch() in string.whitespace)):
+        if curch() == '%':
+            skip_comment()
+        nextch()
+
+# skip whitespace and check if the next char on the input is in s
+# if so, return it, and advance the character pointer.
+# If not or at EOF, return None
+def expect(s):
+    skip_whitespace()
+    c = curch()
+    if c == None:
+        return None
+    if c not in s:
+        return None
+    else:
+        nextch()
+        return c
+
+# skip whitespace and check if the next char on the input is in s
+# if so, return it, and advance the character pointer.
[…]

### 2012-10-29 bin/pybcompact.py

@@ -1,0 +1,120 @@
+#! /usr/bin/env python
+# Original code by Adriaan de Groot. 
+#
+# Disributed under the BSD license
+#
+# Extract only those entries from a set of bibtex files that are cited in
+# a latex document, and write those to standard output.
+#
+# usage: pybcompact  <latexfile> <bibtexfiles...>
+
+import sys, string, os, fnmatch
+import pybaux,pybbib
+
+def usage ():
+    print "usage: pybcompact <latexfile> <bibtoplevel>"
+    return
+
+def error (msg):
+    sys.stderr.write("\npybcompact: error: %s\n" % msg)
+    sys.exit (1)
+    return
+
+def warning (msg):
+    sys.stderr.write("pybcompact: warning: %s\n" % msg)
+    return
+
+def collate (bibs,cites):
+    found = []
+    crossref = {}
+    for c in cites:
+        if bibs.has_key(c):
+            useentry = merged_bibs[c]
+            found.append(c)
+            if useentry.has_key("crossref"):
+                xref = useentry["crossref"]
+                xref = xref.strip()
+                xref = xref.strip("{}")
+                xref = xref.strip("}{")
+                crossref[xref] = 1
+        else:
+            warning("Not found: %s" % (c))
+    return ( found,crossref.keys() )
+
+# test input arguments
+if len (sys.argv) < 3:
+    usage ()
+    sys.exit (1)
+
+# get arguments
+latexfile  = sys.argv [1]
+bibdir = sys.argv [2]
+
+# obtain citation keys from latex aux file
+try:
+    citations = pybaux.aux_citations(latexfile)
+except pybaux.AuxError, err:
+    error ('%s' % err)
+
+# ensure citation keys unicity
+tmp = {}
+for c in citations:
+    tmp [c] = 1
+citations = tmp.keys ()
+
+# is there something to do ?
+if len (citations) == 0:
+    sys.stderr.write("\npybcompact: no entries in aux file\n")
+    sys.exit(0)
+
+# load all bib files and merge in one big dictionary
+merged_bibs = {}
+bibfiles = []
+pattern = '*.bib'
+
+for root, dirs, files in os.walk(bibdir):
+    for filename in fnmatch.filter(files, pattern):
+        bibfiles.append(os.path.join(root, filename))
+
+for bib in bibfiles:
+    newbib = pybbib.bib_parse(bib,sys.stderr)
+    for k in newbib.keys():
+        merged_bibs[k] = newbib[k]
+
+#
[…]

### 2012-10-29 latex/paper.tex
More content: and bibliography

@@
 
-\author{\IEEEauthorblockN{Georgios Gousios}
+\author{\IEEEauthorblockN{Erik Meijer}
+\and
+\IEEEauthorblockN{Georgios Gousios}
 \IEEEauthorblockA{
@@
 Delft, The Netherlands\\
-Email: G.Gousios@tudelft.nl} 
+Email: {H.J.M.Meijer, G.Gousios, Arie.vanDeursen@tudelft.nl} 
+}  
 \and
-\IEEEauthorblockN{Erik Meijer}
-\IEEEauthorblockA{Microsoft \\
-Rendmond, WA, USA\\
-Email: H.J.M.Meijer@tudelft.nl} 
+\IEEEauthorblockN{Arie van Deursen}
 }
@@
 
-  In the last few years, the popularity of functional programming as a way
-  of solving computational problems has increased significantly. While most
-  computer science curricula do include a course on functional programming,
-  in many cases it is disconnected from practical applications, which is
+  In the last few years, the popularity of functional programming as a way of
+  solving computational problems has increased significantly. While most
+  computer science curricula do include a course on functional programming, in
+  many cases it is disconnected from practical applications, which is
   precisely where functional programming shines. To fill in this gap, we
-  designed a functional programming course that demanded from students to learn
-  by experience with real world applications. In this work, we present our
-  experiences with conducting this course.
+  designed a functional programming course that demanded from students to
+  learn by experience with real world applications. In this work, we present
+  our experiences with conducting this course.
+
 \end{abstract}
@@
 
-Due to a variety of reasons, including the advent of cloud computing, the rising
-rate of information production and the necessity to reach the market fast,
-currently, large corporations and start-ups alike are investigating
-alternative programming and information storage models. As a result, during the
-last few years, the practical software engineering field is witnessing a noticeable shift towards functional programming. Scripting languages,
-notably Javascript and Ruby, pioneered the introduction of functional
-concepts, such as closures and lambda functions, to mainstream programming. A
-new wave of programming languages, developed to overcome the expressiveness and
+Due to a variety of reasons, including the advent of cloud computing, the
+rising rate of information production and the necessity to reach the market
+fast, currently, large corporations and start-ups alike are investigating
+alternative programming and information storage models. As a result, during
+the last few years, the practical software engineering field is witnessing a
+noticeable shift towards functional programming. Scripting languages, notably
+Javascript and Ruby, pioneered the introduction of functional concepts, such as
+closures and lambda functions, to mainstream programming. A new wave of
+programming languages, developed to overcome the expressiveness and
 complexity limitations exhibited in mainstream languages, have promoted
 functional constructs, such as type safe pattern matching, higher order
-functions and single assignment variables, to first class citizens (Scala). New,
-purely functional, languages have emerged to fill in the remaining gaps (F\#,
-Clojure), often introducing significant advancements in their field of
-specialisation (such as Erlang in distributed fault-tolerant systems). Finally,
-large scale information processing systems such as Map-Reduce and domain
-specific languages such as LINQ have integrated functional concepts to ease the
-expression of computations.
+functions and single assignment variables, to first class citizens (Scala).
+New, purely functional, languages have emerged to fill in the remaining gaps
+(F\#, Clojure), often introducing significant advancements in their field of
+specialisation (such as Erlang in distributed fault-tolerant systems).
+Finally, large scale information processing systems such as Map-Reduce and
+domain specific languages such as LINQ have integrated functional concepts to
+ease the expression of computations.
 
@@
 While functional programming has been taught for long in computer science
-departments~\cite{Joost93}, curricula tend to emphasize functional programming theory rather
-than practical applications. Special programming languages are used to teach
-functional programming specific concepts, while little connection is made to how
-those concepts can be transfered to non-functional languages.  The course at TU
-Delft attempted to teach functional programming theory in the classroom and then
-expect students to apply those concepts in the (non-purely functional) language
-of their choice. The course had a very strong teaching by example focus:
-students were expected to participate in both in-classroom exercises, homework
-assignments and implement a real world system as a final project.
+departments~\cite{Joost93}, curricula tend to emphasize functional programming
+theory rather than practical applications. Special programming languages are
+used to teach functional programming specific concepts, while little connection
+is made to how those concepts can be transfered to non-functional languages.
+The course at TU Delft attempted to teach functional programming theory in the
+classroom and then expect students to apply those concepts in the (non-purely
+functional) language of their choice. The course had a very strong teaching by
+example focus: students were expected to participate in both in-classroom
+exercises, homework assignments and implement a real world system as a final
+project.
+
+\section{Challenges}
+
+While planning the course, we dealt with the following challenges related
+to the course's organization and content.
+
+\subsection{Heterogeneous Background of participants}
+
+The functional programming course is elective at the MSc curriculum. Students
+from all departments in the Electrical Engineering, Mathematics and Computer
+science faculty are allowed to participate. For practical reasons, the
+enrollment to the course was limited to 15 participants. This permitted
+us to get to know the students individually and to provide them with
+personalized support.
+
+From the students that enrolled, most have received formal introduction to
+imperative and object oriented programming in their bachelors curriculum, while
+through participation to other courses they had limited exposure to functional
+programming (i.e. Map/Reduce~\cite{Dean04} in the web information systems
+course). Two of the students were majoring in computer engineering, which meant
+that their programming experience was restricted. Several students also had work
+experience as programmers as part of their industrial placement or through
+participation to start up ventures.
+
+The expected diverse backgrounds of the students guaranteed that no generic
+introduction to the topic would be sufficient. While the
+
+\subsection{Choosing the appropriate topics}
+
+Functional programming is arguably the oldest programming paradigm. In it is
+purest form, it is based on a minimalistic theoretical background
+($\lambda$-calculus~\cite{Baren84}). Relatively recent
+formulations~\cite{Meije91, Wadle93} also introduced concepts from category
+theory. While the every day use of functional programming does not necessarily
+require the programmer to be aware of the theoretical background, understanding
+it usually leads to more elegant algorithmic solutions. Functional
+programming teaching is also associated with advanced type systems; the flagship
+functional programming languages (Haskell and the {\sc ml} family) both offer
+very sophisticated support for types. Consequently, teaching the full spectrum
+of functional programming theory and techniques would have been impossible in a
+half-semester course; instead we decided to focus on practical aspects of data
+processing, transformations and state representation using functional techniques.
+
+\subsection{Choosing the appropriate programming language}
 
@@
   \item To introduce students to basic functional programming concepts, such
-    as higher order functions, monads and type inference.
+    as higher order functions, monads and advanced type systems.
 
-  \item To introduce students to solving data processing problems in a 
-    functional way.
+  \item To introduce students to the approach of viewing data processing
+    problems as a series of function applications.
 
@@
 
-The course consisted of a series of lectures
-
-\subsection{Student profile}
-
-The course was elective at the MSc curriculum. Students from all departments in
-the Electrical Engineering, Mathematics and Computer science faculty were
-allowed to participate. Most students had received formal introduction to
-imperative and object oriented programming in their bachelors curriculum, while
-through participation to other courses they had limited exposure to functional
-programming (i.e. Map/Reduce in the web information systems course). Two of the
-students were majoring in computer engineering, which meant that their
-programming experience was somewhat limited. Several students also had work
-experience as programmers as part of their industrial placement or through
-participation to start up ventures.
-
-For organizational reasons, the participation to the course was limited to 15
-students.
+The course consisted of a series of lectures, of which two where invited
+lectures by an external instructor, laboratories, a series of intermediate
+assignments and a final assignment. The students would be evaluated by their
+performance on the final assignment only.
 
 \subsection{Lectures}
+
+\cite{Hutto07}
 
@@
   \item Multi-source real time data processing. Students worked on a 
-    programming language popularity index
+    programming language popularity index, based on data from the Github
+    event stream and the StackOverflow tag cloud {\sc api}.
 
   \item Implementation of Haskell constructs in Javascript. Students implemented
-    the Haskell prelude (basic functions for list manipulation).
+    the Haskell Prelude (basic functions for list manipulation).
 
@@
 
+  \item Implementation of simple constraint solvers.
+
   \item Machine learning algorithms. Students implemented Na\"ive Bayes 
-    classifiers and K-means clusters.
+    classifiers and K-means clustering algorithms.
 
 \end{itemize}
+
+The assignments were open ended, as there was no fixed set of targets for
+each one of them. The deliverables for the assignments were a repository
+with the source code and short report describing the solution that the
+students came up with. The students were also required to have a working
+demo by the end of the course time.
+
+\subsection{The labs}
 
@@
 
+Teaching the concepts of solving problems algorithmically can be a daunting
+task~\cite{Futsc06}; especially more so, when the subjects are already in a
+specific mindset. The assumption of the presented course was that imperative and
+functional programming are really the two sides of the same coin; by focusing on
+how students can apply rather than just be taught concepts of functional
+programming using the, typically imperative, languages they already know, allows
+them to better appreciate the strong points and weaknesses of each paradigm. In
+our experience, the approach has been successful. Both throughout the initial
+homework assignments and in their projects, the students demonstrated a high
+degree of appreciation of functional programming concepts. Above all, we
+believe that it was the hands-on approach through tough homework exercises
+employed in this course that allowed students to sharpen their skills and
+understand the concepts that were taught during the course.
+
 \section*{Acknowledgements}
 
-We would like to thank\ldots
+We would like to thank the course participants for their enthusiasm and
+dedication.
 

### 2012-10-30 latex/paper.tex
Teapot screenshots

@@
 \usepackage{times}
+\usepackage{subfigure}
 
@@
 
+\begin{figure*}
+\centering
+\subfigure[Reference, rendered with Java graphics]{
+\includegraphics[scale=0.173]{teapot-reference.png}
+\label{fig:teapot-reference}
+}
+\subfigure[Decomposition in Scala, rendered with Java graphics]{
+\includegraphics[scale=0.17]{teapot-java-canvas.png}
+\label{fig:teapot-scala}
+}
+\subfigure[Decomposition in Scheme, rendered with appropriately positioned {\sc html} \textsf{div}
+elements]{
+\includegraphics[scale=0.2]{teapot-html-divs.png}
+\label{fig:teapot-html-divs}
+}
+\subfigure[Decomposition in Javascript, rendered with {\sc html5} graphics]{
+\includegraphics[scale=0.31]{teapot-html-canvas.png}
+\label{fig:teapot-html-canvas}
+}
+\label{fig:proj}
+\caption[]{Example results from the teapot experiment.}
+\end{figure*}
+
 The students' response to the exercise was overwhelming. For four days,
@@
 
-\section*{Acknowledgements}
+\section*{Acknowledgements and Availability}
 

### 2012-10-30 latex/paper.tex
Choosing the programming language

@@
 \usepackage{subfigure}
+\usepackage{url}
 
@@
 
-\subsection{Heterogeneous Background of participants}
+\subsection{Heterogeneous background of participants}
 
@@
 The expected diverse backgrounds of the students guaranteed that no generic
-introduction to the topic would be sufficient. While the
+introduction to the topic would be sufficient. For this reason, we decided
+to adopt a hands-on-first approach; the students would have to learn by
+flexing their programming muscles, instead of being gently introduced to 
+the theory through toy examples.
 
@@
 \subsection{Choosing the appropriate programming language}
+
+Programming languages, in addition to enabling the programmer to express a
+series of instructions to be executed by a computer, affect the programmer's
+tough process and consequently her approach towards problem
+solving~\cite{Ivers80}.  When teaching a programming course, it is important to
+be able to demonstrate concepts without interference by the chosen language's
+syntax. However, a non-practical language might have the opposite effect; our
+experience has shown that the further a demonstration language is from practical
+application, the less important student feel the taught concepts are.
+Fortunately, most functional programming concepts can be expressed cleanly by
+several widespread languages: for example, Javascript, Ruby and {\sc c\#} have
+closures, first class and higher order functions, while a number of emerging
+languages, such as Scala, {\sc f\#} and Clojure are functional languages working
+on familiar development platforms. Apart from staying compatible with the
+existing literature, there is no practical reason that necessitates teaching of
+functional programming strictly in a language like Haskell or {\sc ml}.
+
+The above led us to not choose any particular language for the course. 
+During the lectures, the demonstration languages would be Haskell and
+{\sc c\#}. We actively encouraged students to use the language of their
+choice on the platform of their choice to carry out homework assignments
+and the course's final project.
 
@@
 core, the exercise required students to decompose arbitrary triangles, which
-comprised the input Utah Teapot model, to a series right triangles. The
-students had to come up with the decomposition method (using some form of
-analytic geometry), a decomposition restriction criterion to stop the
-decomposition when triangles could not be rendered on screen and a method to
-recursively apply the above mentioned transformations on the input data. As
-always the students could decide the implementation language of their choice.
+comprised the input Utah Teapot model, to a series right triangles. The students
+had to come up with the decomposition method (using some form of analytic
+geometry), a decomposition restriction criterion to stop the decomposition when
+triangles could not be rendered on screen and a method to recursively apply the
+above mentioned transformations on the input data. As always, the students could
+decide the implementation language of their choice. The exercise was given as a
+mid-week homework assignment, between the Monday and the Friday lectures.
 
@@
 \centering
-\subfigure[Reference, rendered with Java graphics]{
+\label{fig:teapots}
+\subfigure[Input drawing, rendered with Java graphics]{
 \includegraphics[scale=0.173]{teapot-reference.png}
@@
 }
-\label{fig:proj}
-\caption[]{Example results from the teapot experiment.}
+\caption[]{Example results from the teapot exercise.}
 \end{figure*}
 
-The students' response to the exercise was overwhelming. For four days,
-the students worked hard to
+The student's response was overwhelming. Even though it was made clear that
+the exercise would not contribute to the final grade, the will to apply the
+data processing techniques taught during the lectures motivated
+the students to work very hard. The fact that they were instructed to use
+their favourite language enabled them to focus on decomposing the problem 
+in a series of testable calculation steps rather than meddling with the
+intricacies of a pure functional language. Solutions where provided in
+Scala, {\sc c\#}, Scheme, Javascript and Haskell. 
+Example renderings produced by student programs can be seen in 
+Figure~\ref{fig:teapots}.
 
@@
 We would like to thank the course participants for their enthusiasm and
-dedication.
+dedication. The source code to all student projects and the teapot
+exercise can be found at (or linked from) \url{https://github.com/gousiosg/teapots}.
 

### 2012-10-30 latex/paper.tex
More content + refs

@@
 Delft, The Netherlands\\
-Email: {H.J.M.Meijer, G.Gousios, Arie.vanDeursen@tudelft.nl} 
+Email: {H.J.M.Meijer, G.Gousios@tudelft.nl} 
 }  
-\and
-\IEEEauthorblockN{Arie van Deursen}
 }
@@
 
-The aim of the course is to teach the principles of pure functional
-programming, and the corresponding Category theoretical principles, using the
-Haskell programming language. More specifically, the educational purposes
-of the course were:
+The aim of the course was to teach the principles of pure functional
+programming, and the corresponding Category theoretical principles. 
+More specifically, the educational purposes of the course were:
 
@@
 
+  \item To
+
 \end{itemize}
@@
 The course consisted of a series of lectures, of which two where invited
-lectures by an external instructor, laboratories, a series of intermediate
-assignments and a final assignment. The students would be evaluated by their
-performance on the final assignment only.
+lectures by an external instructor, and laboratory sessions. In total, the
+course consisted of 14 hours of lectures and 8 hours of lab work.
+The students
+also had to carry out a series of intermediate assignments and a final project.
+The students would be evaluated by their performance on the final assignment. 
+The lectures took place in a quarter (half semester), during 3 intense 
+weeks; the students had another 6 weeks to finish their final projects.
 
@@
 
-\cite{Hutto07}
+The lectures covered the following topics:
+
+\begin{itemize}
+
+  \item Functions and functional composition, higher order functions, recursive
+    functions, avoiding recursion
+
+  \item Basic types, function types, currying
+
+  \item Lists, mapping and traversal using folds
+
+  \item Monads, composition and their application on the {\sc linq} query 
+    language~\cite{Meije11}
+
+  \item Monoids, functors and their application on data structure processing
+
+  \item Functional formalization of the Map/Reduce data processing
+    paradigm~\cite{Lamme08}
+
+\end{itemize}
+
+The lecture material was loosely based on Graham Hutton's functional programming
+course at the University of Nottingham. The course's recommended book was
+Programming in Haskell by the same author~\cite{Hutto07}.
 
@@
 
-\subsection{The labs}
+\subsection{Labs}
+
+The purpose of the lab sessions where to help students carry out their final
+assignments and explain better concepts that might not be clear to students. 
+Students convened every week, presented their progress and discussed with 
+each other and with the lab demonstrator design and practical issues that
+arised in the implementation of their projects. To further motivate students
+to work on their projects, one of the lab sessions was converted to a full
+day coding sprint.
 
@@
 
-\subsection{Teaching Monads}
+\subsection{Teaching by example}
 
 \subsection{The teapot experiment}
+
+\begin{figure*}
+\centering
+\label{fig:teapots}
+\subfigure[Input drawing, rendered with Java graphics]{
+\includegraphics[scale=0.173]{teapot-reference.png}
+\label{fig:teapot-reference}
+}
+\subfigure[Decomposition in Scala, rendered with Java graphics]{
+\includegraphics[scale=0.17]{teapot-java-canvas.png}
+\label{fig:teapot-scala}
+}
+\subfigure[Decomposition in Scheme, rendered with appropriately positioned {\sc html} \textsf{div}
+elements]{
+\includegraphics[scale=0.2]{teapot-html-divs.png}
+\label{fig:teapot-html-divs}
+}
+\subfigure[Decomposition in Javascript, rendered with {\sc html5} graphics]{
+\includegraphics[scale=0.31]{teapot-html-canvas.png}
+\label{fig:teapot-html-canvas}
+}
+\caption[]{Example results from the teapot exercise.}
+\end{figure*}
 
@@
 
-\begin{figure*}
-\centering
-\label{fig:teapots}
-\subfigure[Input drawing, rendered with Java graphics]{
-\includegraphics[scale=0.173]{teapot-reference.png}
-\label{fig:teapot-reference}
-}
-\subfigure[Decomposition in Scala, rendered with Java graphics]{
-\includegraphics[scale=0.17]{teapot-java-canvas.png}
-\label{fig:teapot-scala}
-}
-\subfigure[Decomposition in Scheme, rendered with appropriately positioned {\sc html} \textsf{div}
-elements]{
-\includegraphics[scale=0.2]{teapot-html-divs.png}
-\label{fig:teapot-html-divs}
-}
-\subfigure[Decomposition in Javascript, rendered with {\sc html5} graphics]{
-\includegraphics[scale=0.31]{teapot-html-canvas.png}
-\label{fig:teapot-html-canvas}
-}
-\caption[]{Example results from the teapot exercise.}
-\end{figure*}
-
 The student's response was overwhelming. Even though it was made clear that
@@
 \subsection{Teaching Map/Reduce}
+
+\subsection{The coding sprint day}
+
+Coding sprints or ``hackathons'' have long been used by open source software
+projects to speed up development and increase participation of community
+members~\cite{Adams11}. In a typical sprint, project members set measurable
+targets and work intensively to
 

### 2012-11-01 latex/paper.tex
Fix paper formatting

@@
 
-\author{\IEEEauthorblockN{Erik Meijer}
-\and
-\IEEEauthorblockN{Georgios Gousios}
+\author{\IEEEauthorblockN{Erik Meijer and Georgios Gousios}
 \IEEEauthorblockA{
@@
 Delft, The Netherlands\\
-Email: {H.J.M.Meijer, G.Gousios@tudelft.nl} 
+Email: \{h.j.m.meijer, g.gousios\}@tudelft.nl
 }
@@
   precisely where functional programming shines. To fill in this gap, we
-  designed a functional programming course that demanded from students to
-  learn by experience with real world applications. In this work, we present
-  our experiences with conducting this course.
+  designed a functional programming course that  required students to
+  learn by experience with real world applications. In this paper, we present
+  the course's design and outline our experience from delivering it.
 
 \end{abstract}
+
+\begin{IEEEkeywords}
+functional programming, teaching
+\end{IEEEkeywords}
 

### 2012-11-01 latex/paper.tex
Fill in text here and there

@@
 is made to how those concepts can be transfered to non-functional languages.
-The course at TU Delft attempted to teach functional programming theory in the
+The course at TU Delft attempted to teach functional programming principles in the
 classroom and then expect students to apply those concepts in the (non-purely
@@
 with the source code and short report describing the solution that the
-students came up with. The students were also required to have a working
+students came up with. The students were also required to present a working
 demo by the end of the course time.
@@
 \subsection{Teaching by example}
+
+The focus of teaching was to present functional programming from a 
+practical aspect.
+
+The proverbial ``six ways to understand monads'' included presentations of the
+formal theoretical concepts, visualizations of monads as blenders (based on the
+monadic property that once a value is captured, it is then contained),
+applications in Haskell for encapsulating state (through the \textsf{Maybe}
+monad), applications to generic programming in {\sc c\#}, applications to {\sc
+linq} and the reactive extensions framework, and, finally, a whiteboard exercise
+where students where prompted to participate to the specification of Haskell's
+\textsf{IO} monad from basic principles.
 
@@
 
-The student's response was overwhelming. Even though it was made clear that
-the exercise would not contribute to the final grade, the will to apply the
-data processing techniques taught during the lectures motivated
-the students to work very hard. The fact that they were instructed to use
-their favourite language enabled them to focus on decomposing the problem 
-in a series of testable calculation steps rather than meddling with the
-intricacies of a pure functional language. Solutions where provided in
-Scala, {\sc c\#}, Scheme, Javascript and Haskell. 
-Example renderings produced by student programs can be seen in 
-Figure~\ref{fig:teapots}.
+The student's response was overwhelming. Even though it was made clear that the
+exercise would not contribute to the final grade, the will to apply the data
+processing techniques taught during the lectures motivated the students to work
+very hard. The fact that they were instructed to use their favourite language
+enabled them to focus on decomposing the problem in a series of testable
+calculation steps rather than meddling with the intricacies of a pure functional
+language.  Despite the advice, students were eager to test-drive the newly
+acquired knowledge in a functional language; apart from {\sc c\#} and
+Javascript, solutions were also provided in Scala, Scheme and Haskell. Example
+renderings produced by student programs can be seen in Figure~\ref{fig:teapots}.
 
 \subsection{Teaching Map/Reduce}
+
+According to many students, the Map/Reduce data analysis framework was what
+drove them to attend the course. During the course, several examples
+where presented
+
+The result was that in their projects, students used implicit, through
+the selected language's library collection methods, or explicit, by
+formalizing data processing as a series of in memory Map/Reduce operations, forms
+of Map/Reduce to process the data.
+
+Out of personal interest, one student even implemented L\"ammel Map/Reduce
+formulation in Scala, and through the language's extension mechanisms,
+adapted it for use as an extension method by any Scala collection type,
+including parallel collections.
 
@@
 members~\cite{Adams11}. In a typical sprint, project members set measurable
-targets and work intensively to
+targets and work intensively in a tight deadline to deliver them. To ensure
+the timely delivery of the student projects, we organized an one day
+coding sprint, during the students where request to i) set an explicit target 
+ii) write code for 7 hours with a short break iii) present a 5 minute 
+demo of their work at the end of the day. All project teams where expected to
+work in the same room, while communication among the teams was facilitated
+by the lab demonstrator, to avoid unecessary distractions.
 
 \subsection{Student Projects}
+
+As expected, the performance of students in project varied.
 

### 2012-11-01 latex/paper.tex
Teaching by example - done

@@
 
-The above led us to not choose any particular language for the course. 
-During the lectures, the demonstration languages would be Haskell and
-{\sc c\#}. We actively encouraged students to use the language of their
-choice on the platform of their choice to carry out homework assignments
+The above led us to not choose any particular language for the course.  During
+the lectures, the demonstration languages where Haskell and {\sc c\#}, while at
+the labs we used Scala. We actively encouraged students to use the language of
+their choice on the platform of their choice to carry out homework assignments
 and the course's final project.
@@
 
-The focus of teaching was to present functional programming from a 
-practical aspect.
+The focus of teaching was to present functional programming from a practical
+aspect; we were mostly interested to teach students what they can do rather than
+the theory behind functional constructs. An important distinction
+that was made early on was that the every day world is imperative; therefore
+while functional programming can be a great tool for thinking about a problem, a
+pure functional solution cannot reflect on the real world. As a consequence it
+is usually best to mix programming paradigms, i.e. imperative or object-oriented
+programming for state representation and functional programming for data
+processing. In our experience, this distinction helped students understand
+that they are already doing functional programming without noticing;
+enforcing data immutability to handle multi-threading problems or passing
+function arguments in scripting languages are applications of functional
+programming principles. The moment students realized this fact, their
+attention was captured to the lectures.
+
+During both lectures and labs, all non-whiteboard examples where actually
+entered in a real programming environment (LinqPad or the Scala command-line
+read-evaluate loop) and the results of the evaluation where discussed with the
+students. Participation was encouraged by modifying the examples and asking the
+students to perform the evaluation, before the example was executed. The
+examples varied from list manipulation with higher-order functions to 
+a toy actor model example.
 
@@
 monad), applications to generic programming in {\sc c\#}, applications to {\sc
-linq} and the reactive extensions framework, and, finally, a whiteboard exercise
-where students where prompted to participate to the specification of Haskell's
-\textsf{IO} monad from basic principles.
+linq} and the reactive extensions frameworks, and, finally, a whiteboard
+exercise where students were prompted to participate in the ad-hoc specification
+of Haskell's \textsf{IO} monad from basic principles. Once again, the examples
+enabled the students to understand the composition and state containment
+properties of an otherwise difficult to comprehend theoretical concept.
 
-\subsection{The teapot experiment}
+\subsection{The teapot exercise}
 
@@
 One of the highlights of the taught period of the course was the teapot
-experiment. While not a formal experiment in the scientific sense of the word,
+exercise, which 
 we used it to draw the student's attention to the following facts:
@@
 calculation steps rather than meddling with the intricacies of a pure functional
-language.  Despite the advice, students were eager to test-drive the newly
+language. In spite of the advice, students were eager to test-drive the newly
 acquired knowledge in a functional language; apart from {\sc c\#} and
@@
 
-\subsection{Teaching Map/Reduce}
+\subsection{Map/Reduce}
 
 According to many students, the Map/Reduce data analysis framework was what
-drove them to attend the course. During the course, several examples
-where presented
-
-The result was that in their projects, students used implicit, through
-the selected language's library collection methods, or explicit, by
-formalizing data processing as a series of in memory Map/Reduce operations, forms
-of Map/Reduce to process the data.
-
+drove them to attend the course. The lecture on Map/Reduce featured an invited
+talk by Ralf L\"ammel, who presented his formulation of Map/Reduce using
+functional programming principles~\cite{Lamme08} and explained how those can be
+applied to process semi-structured data. The talk resonated to students; in
+their projects, students used implicit, through the selected language's library
+collection methods, or explicit, by formalizing data processing as a series of
+in memory Map/Reduce operations, forms of Map/Reduce to process their datasets.
 Out of personal interest, one student even implemented L\"ammel Map/Reduce
-formulation in Scala, and through the language's extension mechanisms,
-adapted it for use as an extension method by any Scala collection type,
-including parallel collections.
+formulation in Scala, and through the language's extension mechanisms, adapted
+it for use as an extension method by any Scala collection type, including
+parallel collections.
 
@@
 work in the same room, while communication among the teams was facilitated
-by the lab demonstrator, to avoid unecessary distractions.
+by the lab demonstrator, to avoid unnecessary distractions.
 
-\subsection{Student Projects}
+The sprint helped students kick start their projects.
 
-As expected, the performance of students in project varied.
+\subsection{The student projects}
+
+As expected, the performance of students in their projects varied.
 
@@
 dedication. The source code to all student projects and the teapot
-exercise can be found at (or linked from) \url{https://github.com/gousiosg/teapots}.
+exercise can be found at (or linked from) \url{http://swerl.tudelft.nl/bin/view/Main/FunctionalProgrammingCourse}.
 

### 2012-11-01 latex/paper.tex
Final version

@@
 \IEEEauthorblockA{
+Software Engineering Research Group\\
 Delft University of Technology\\
@@
 
-Due to a variety of reasons, including the advent of cloud computing, the
-rising rate of information production and the necessity to reach the market
-fast, currently, large corporations and start-ups alike are investigating
-alternative programming and information storage models. As a result, during
-the last few years, the practical software engineering field is witnessing a
-noticeable shift towards functional programming. Scripting languages, notably
-Javascript and Ruby, pioneered the introduction of functional concepts, such as
-closures and lambda functions, to mainstream programming. A new wave of
-programming languages, developed to overcome the expressiveness and
-complexity limitations exhibited in mainstream languages, have promoted
-functional constructs, such as type safe pattern matching, higher order
-functions and single assignment variables, to first class citizens (Scala).
-New, purely functional, languages have emerged to fill in the remaining gaps
-(F\#, Clojure), often introducing significant advancements in their field of
-specialisation (such as Erlang in distributed fault-tolerant systems).
-Finally, large scale information processing systems such as Map-Reduce and
-domain specific languages such as LINQ have integrated functional concepts to
-ease the expression of computations.
+Due to a variety of reasons, including the advent of cloud computing, the rising
+rate of information production and the necessity to reach the market fast,
+currently, large corporations and start-ups alike are investigating alternative
+programming and information storage models. As a result, during the last few
+years, the practical software engineering field is witnessing a noticeable shift
+towards functional programming. Scripting languages, notably Javascript and
+Ruby, pioneered the introduction of functional concepts, such as closures and
+lambda functions, to mainstream programming. A new wave of programming
+languages, developed to overcome the expressiveness and complexity limitations
+exhibited in mainstream languages, have promoted functional constructs, such as
+type safe pattern matching, higher order functions and single assignment
+variables, to first class citizens (Scala).  New, purely functional, languages
+have emerged to fill in the remaining gaps (F\#, Clojure), often introducing
+significant advancements in their field of specialisation (such as Erlang in
+distributed fault-tolerant systems).  Finally, large scale information
+processing systems such as Map/Reduce~\cite{Dean04} and domain specific
+languages such as {\sc linq} have integrated functional concepts to ease the
+expression of computations.
 
@@
 
-The functional programming course is elective at the MSc curriculum. Students
+The functional programming course is elective at the MSc level. Students
 from all departments in the Electrical Engineering, Mathematics and Computer
@@
 through participation to other courses they had limited exposure to functional
-programming (i.e. Map/Reduce~\cite{Dean04} in the web information systems
+programming (i.e. Map/Reduce in the web information systems
 course). Two of the students were majoring in computer engineering, which meant
@@
 
-The expected diverse backgrounds of the students guaranteed that no generic
+The diverse backgrounds of the students guaranteed that no generic
 introduction to the topic would be sufficient. For this reason, we decided
 to adopt a hands-on-first approach; the students would have to learn by
-flexing their programming muscles, instead of being gently introduced to 
+flexing their programming muscles, instead of being introduced to 
 the theory through toy examples.
@@
 theory. While the every day use of functional programming does not necessarily
-require the programmer to be aware of the theoretical background, understanding
-it usually leads to more elegant algorithmic solutions. Functional
-programming teaching is also associated with advanced type systems; the flagship
-functional programming languages (Haskell and the {\sc ml} family) both offer
-very sophisticated support for types. Consequently, teaching the full spectrum
-of functional programming theory and techniques would have been impossible in a
+require the programmer to be aware of the theory, understanding it usually leads
+to more elegant algorithmic solutions. Functional programming teaching is also
+associated with advanced type systems; indeed, the flagship functional
+programming languages (Haskell and the {\sc ml} family) both offer very
+sophisticated support for types. Consequently, teaching the full spectrum of
+functional programming theory and techniques would have been impossible in a
 half-semester course; instead we decided to focus on practical aspects of data
-processing, transformations and state representation using functional techniques.
+processing, transformations and state representation using functional
+techniques.
 
@@
 series of instructions to be executed by a computer, affect the programmer's
-tough process and consequently her approach towards problem
+thought process and consequently her approach towards problem
 solving~\cite{Ivers80}.  When teaching a programming course, it is important to
@@
 closures, first class and higher order functions, while a number of emerging
-languages, such as Scala, {\sc f\#} and Clojure are functional languages working
-on familiar development platforms. Apart from staying compatible with the
-existing literature, there is no practical reason that necessitates teaching of
-functional programming strictly in a language like Haskell or {\sc ml}.
+languages, such as Scala, {\sc f\#} and Clojure are functional languages
+implemented on familiar development platforms. Apart from staying compatible
+with the existing literature, there is no practical reason that necessitates
+teaching of functional programming strictly in a language like Haskell or {\sc
+ml}.
 
@@
 
-The aim of the course was to teach the principles of pure functional
+The high level goal of the course was to teach the principles of functional
 programming, and the corresponding Category theoretical principles. 
@@
 
-  \item To introduce students to basic functional programming concepts, such
-    as higher order functions, monads and advanced type systems.
+  \item To introduce students to basic functional programming concepts,
+    higher order functions, monads and advanced type systems.
 
-  \item To introduce students to the approach of viewing data processing
+  \item To introduce students to the approach of expressing data processing
     problems as a series of function applications.
@@
     functional environments.
-
-  \item To
 
@@
 The students
-also had to carry out a series of intermediate assignments and a final project.
+also had to carry out a series of homework assignments and a final project.
 The students would be evaluated by their performance on the final assignment. 
@@
 
-The assignments were open ended, as there was no fixed set of targets for
-each one of them. The deliverables for the assignments were a repository
+To counterbalance the differences in student experience level, 
+the assignments were open ended; students could drive their projects as
+far as they where willing to.
+The deliverables for the assignments were a repository
 with the source code and short report describing the solution that the
@@
 The purpose of the lab sessions where to help students carry out their final
-assignments and explain better concepts that might not be clear to students. 
+assignments and further explain concepts that might not be clear to students. 
 Students convened every week, presented their progress and discussed with 
@@
 aspect; we were mostly interested to teach students what they can do rather than
-the theory behind functional constructs. An important distinction
-that was made early on was that the every day world is imperative; therefore
-while functional programming can be a great tool for thinking about a problem, a
-pure functional solution cannot reflect on the real world. As a consequence it
-is usually best to mix programming paradigms, i.e. imperative or object-oriented
-programming for state representation and functional programming for data
-processing. In our experience, this distinction helped students understand
-that they are already doing functional programming without noticing;
-enforcing data immutability to handle multi-threading problems or passing
-function arguments in scripting languages are applications of functional
-programming principles. The moment students realized this fact, their
-attention was captured to the lectures.
+the theory behind functional constructs. An important distinction that was made
+early on was that the world is imperative; therefore while functional
+programming can be a great tool for thinking about a problem, a pure functional
+solutions cannot reflect on the real world. As a consequence, it is usually best
+to mix programming paradigms, i.e. imperative or object-oriented programming for
+state representation and functional programming for data processing. In our
+experience, this distinction helped students understand that they are already
+doing functional programming without noticing; enforcing data immutability to
+handle multi-threading problems or passing function arguments in scripting
+languages are applications of functional programming principles. The moment
+students realized this fact, their attention was captured to the lectures.
 
 During both lectures and labs, all non-whiteboard examples where actually
-entered in a real programming environment (LinqPad or the Scala command-line
+entered in a real programming environment (LinqPad or the Scala command line
 read-evaluate loop) and the results of the evaluation where discussed with the
@@
 elements]{
-\includegraphics[scale=0.2]{teapot-html-divs.png}
+\includegraphics[scale=0.195]{teapot-html-divs.png}
 \label{fig:teapot-html-divs}
@@
 \subfigure[Decomposition in Javascript, rendered with {\sc html5} graphics]{
-\includegraphics[scale=0.31]{teapot-html-canvas.png}
+\includegraphics[scale=0.295]{teapot-html-canvas.png}
 \label{fig:teapot-html-canvas}
@@
 \caption[]{Example results from the teapot exercise.}
+
+\vspace{-1.5em}
 \end{figure*}
@@
 One of the highlights of the taught period of the course was the teapot
-exercise, which 
-we used it to draw the student's attention to the following facts:
+exercise, which we used it to draw the student's attention to the following
+facts:
 
@@
 comprised the input Utah Teapot model, to a series right triangles. The students
-had to come up with the decomposition method (using some form of analytic
-geometry), a decomposition restriction criterion to stop the decomposition when
-triangles could not be rendered on screen and a method to recursively apply the
-above mentioned transformations on the input data. As always, the students could
-decide the implementation language of their choice. The exercise was given as a
-mid-week homework assignment, between the Monday and the Friday lectures.
+had to come up with the decomposition method (using an analytic geometry
+method), a decomposition termination criterion to stop the decomposition when
+triangles are too small to be rendered on screen and a method to recursively
+apply the above mentioned transformations on the input data. As always, the
+students could decide the implementation language of their choice. The exercise
+was given as a mid-week homework assignment, between the Monday and the Friday
+lectures.
 
@@
 enabled them to focus on decomposing the problem in a series of testable
-calculation steps rather than meddling with the intricacies of a pure functional
+calculation steps rather than meddling with the intricacies of learning a new 
 language. In spite of the advice, students were eager to test-drive the newly
-acquired knowledge in a functional language; apart from {\sc c\#} and
+acquired knowledge using a functional language; apart from {\sc c\#} and
 Javascript, solutions were also provided in Scala, Scheme and Haskell. Example
@@
 drove them to attend the course. The lecture on Map/Reduce featured an invited
-talk by Ralf L\"ammel, who presented his formulation of Map/Reduce using
+talk by Prof. Ralf L\"ammel, who presented his formulation of Map/Reduce using
 functional programming principles~\cite{Lamme08} and explained how those can be
@@
 in memory Map/Reduce operations, forms of Map/Reduce to process their datasets.
-Out of personal interest, one student even implemented L\"ammel Map/Reduce
+Out of personal interest, one student even implemented L\"ammel's Map/Reduce
 formulation in Scala, and through the language's extension mechanisms, adapted
@@
 members~\cite{Adams11}. In a typical sprint, project members set measurable
-targets and work intensively in a tight deadline to deliver them. To ensure
-the timely delivery of the student projects, we organized an one day
-coding sprint, during the students where request to i) set an explicit target 
-ii) write code for 7 hours with a short break iii) present a 5 minute 
-demo of their work at the end of the day. All project teams where expected to
-work in the same room, while communication among the teams was facilitated
-by the lab demonstrator, to avoid unnecessary distractions.
+targets and work intensively in a tight deadline to deliver them. To ensure the
+timely delivery of the student projects, we organized an one day coding sprint,
+during which the students were requested to i) set explicit targets ii) write
+code for 7 hours with a short break iii) present a 5 minute demo of their work
+at the end of the day. All project teams where expected to work in the same
+room, while communication among the teams was facilitated by the lab
+demonstrator, to avoid unnecessary distractions.
 
-The sprint helped students kick start their projects.
+Interestingly, the coding sprint was organized on student request. The reason
+was that pressure from other courses did not allow them to be together as teams.
+The sprint taught students the importance of setting realistic development goals
+and the necessity of iterative development. Even though all teams were off
+target at the end of the day, most were able to create a rough prototype in less
+than 7 hours, using functional programming techniques.
 
-\subsection{The student projects}
-
-As expected, the performance of students in their projects varied.
-
+%\subsection{The student projects}
+%
+%For implementing their projects, the students selected languages and frameworks
+%that are currently emerging and popular among non-traditional software
+%development environments. All students did their projects in either Scala or
+%Javascript, with one exception, using frameworks such as Akka (for actors),
+%Scalatra (for implementing {\sc rest} {\sc api}s) or Node.js (for server-side
+%processing). From the implementations it appears that students did apply
+%their 
+%
 \section{Conclusion}
@@
 degree of appreciation of functional programming concepts. Above all, we
-believe that it was the hands-on approach through tough homework exercises
+believe that it was the hands-on approach 
 employed in this course that allowed students to sharpen their skills and
-understand the concepts that were taught during the course.
+understand the concepts that were taught during it.
 
 \section*{Acknowledgements and Availability}
-
 We would like to thank the course participants for their enthusiasm and
 dedication. The source code to all student projects and the teapot
-exercise can be found at (or linked from) \url{http://swerl.tudelft.nl/bin/view/Main/FunctionalProgrammingCourse}.
+exercise can be found at (or linked from) \url{http://swerl.tudelft.nl/bin/view/Main/FunctionalProgrammingCourse}. This work is partially supported by
+Marie Curie {\sc ief} 298930 -- {\sc sefunc}.
 

### 2012-11-01 latex/paper.tex
Erik's corrections

@@
 type safe pattern matching, higher order functions and single assignment
-variables, to first class citizens (Scala).  New, purely functional, languages
-have emerged to fill in the remaining gaps (F\#, Clojure), often introducing
-significant advancements in their field of specialisation (such as Erlang in
-distributed fault-tolerant systems).  Finally, large scale information
-processing systems such as Map/Reduce~\cite{Dean04} and domain specific
-languages such as {\sc linq} have integrated functional concepts to ease the
-expression of computations.
+variables, to first class citizens (Scala, {\sc c\#}).  New, functional
+languages have emerged to fill in the remaining gaps (F\#, Clojure, Racket),
+often introducing significant advancements in their field of specialisation
+(such as Erlang in distributed fault-tolerant systems).  Finally, large scale
+information processing systems such as Map/Reduce~\cite{Dean04} and domain
+specific languages such as {\sc linq} have integrated functional concepts to
+ease the expression of masively parallel computations.
 
@@
 
-Functional programming is arguably the oldest programming paradigm. In it is
+Functional programming is arguably one of the oldest programming paradigms. In it is
 purest form, it is based on a minimalistic theoretical background
@@
 experience has shown that the further a demonstration language is from practical
-application, the less important student feel the taught concepts are.
+application, the less important students feel the taught concepts are.
 Fortunately, most functional programming concepts can be expressed cleanly by
@@
 closures, first class and higher order functions, while a number of emerging
-languages, such as Scala, {\sc f\#} and Clojure are functional languages
+languages, such as Scala, Racket, {\sc f\#} and Clojure are functional languages
 implemented on familiar development platforms. Apart from staying compatible
 with the existing literature, there is no practical reason that necessitates
-teaching of functional programming strictly in a language like Haskell or {\sc
-ml}.
+teaching strictly in a language like Haskell or {\sc ml}.
 
@@
 also had to carry out a series of homework assignments and a final project.
-The students would be evaluated by their performance on the final assignment. 
+The evaluation would be based on student performance on the final assignment. 
 The lectures took place in a quarter (half semester), during 3 intense 
@@
 
-  \item Functions and functional composition, higher order functions, recursive
-    functions, avoiding recursion
+  \item Functions and functional composition, higher order functions, recursion, 
+    avoiding recursion via higher order functions.
 
-  \item Basic types, function types, currying
+  \item Basic types, function types, currying.
 
-  \item Lists, mapping and traversal using folds
+  \item Lists, mapping and traversal using folds.
 
   \item Monads, composition and their application on the {\sc linq} query 
-    language~\cite{Meije11}
+    language~\cite{Meije11}.
 
-  \item Monoids, functors and their application on data structure processing
+  \item Monoids, functors and their application on data structure processing.
 
   \item Functional formalization of the Map/Reduce data processing
-    paradigm~\cite{Lamme08}
+    paradigm~\cite{Lamme08}.
 
@@
 
-  \item Functional programming works best in data transformation scenarios
+  \item Functional programming works best in data transformation scenarios.
+
+  \item Complex problems can be solved by decomposing them into smaller and simpler ones.
 
@@
 The exercise consisted of rendering the Utah Teapot~\cite{Torre06} using any
-graphics primitive of the student's choice using only right triangles.  In its
-core, the exercise required students to decompose arbitrary triangles, which
-comprised the input Utah Teapot model, to a series right triangles. The students
-had to come up with the decomposition method (using an analytic geometry
-method), a decomposition termination criterion to stop the decomposition when
-triangles are too small to be rendered on screen and a method to recursively
+graphics primitive of the student's choice using only right triangles (with one
+horizontal leg).  In its
+core, the exercise required students to decompose a collection of arbitrary triangles, which
+comprised the input Utah Teapot model, to a collection of right triangles. The students
+had to come up with the decomposition method (using some simple analytic
+geometry, this was not a course on 3D graphics), a decomposition termination criterion to stop the decomposition when
+triangles were too small to be rendered on screen and a method to recursively
 apply the above mentioned transformations on the input data. As always, the

### 2012-11-01 latex/paper.tex
Remove references to restrict to 4 pages

@@
 purest form, it is based on a minimalistic theoretical background
-($\lambda$-calculus~\cite{Baren84}). Relatively recent
+($\lambda$-calculus). Relatively recent
 formulations~\cite{Meije91, Wadle93} also introduced concepts from category
@@
 projects to speed up development and increase participation of community
-members~\cite{Adams11}. In a typical sprint, project members set measurable
+members. In a typical sprint, project members set measurable
 targets and work intensively in a tight deadline to deliver them. To ensure the

### 2013-10-22 latex/paper.tex
Update for ICSE 2014

@@
-\documentclass[conference]{IEEEtran}
+\documentclass[conference]{sig-alternate}
 \usepackage{graphicx}
@@
 \usepackage{subfigure}
-\usepackage{url}
+\usepackage{hyperref}
 
@@
 
-\title{Six ways to understand monads\\
-\small{Experiences from the functional programming course at TU Delft}}
+%\author{\IEEEauthorblockN{Erik Meijer and Georgios Gousios}
+%\IEEEauthorblockA{
+%Software Engineering Research Group\\
+%Delft University of Technology\\
+%Delft, The Netherlands\\
+%Email: \{h.j.m.meijer, g.gousios\}@tudelft.nl
+%}
+%}
 
-\author{\IEEEauthorblockN{Erik Meijer and Georgios Gousios}
-\IEEEauthorblockA{
-Software Engineering Research Group\\
-Delft University of Technology\\
-Delft, The Netherlands\\
-Email: \{h.j.m.meijer, g.gousios\}@tudelft.nl
+\title{Teaching Applied Functional Programming\\
+\small{Experiences from the functional programming course at TU Delft}
 }
+
+\numberofauthors{2}
+
+\author{
+\alignauthor
+Erik Meijer\\
+       \affaddr{Delft University of Technology}\\
+       \affaddr{Delft, The Netherlands}\\
+       \email{h.j.m.meijer@tudelft.nl}
+\alignauthor
+Georgios Gousios\\
+       \affaddr{Delft University of Technology}\\
+       \affaddr{Delft, The Netherlands}\\       
+       \email{g.gousios@tudelft.nl}
 }
+
 \maketitle
@@
 
-  In the last few years, the popularity of functional programming as a way of
+  Recently, the popularity of functional programming as a way of
   solving computational problems has increased significantly. While most
@@
   designed a functional programming course that required students to
-  learn by experience with real world applications. In this paper, we present
+  learn by experience in real world applications. In this paper, we present
   the course's design and outline our experience from delivering it.
@@
 
-\begin{IEEEkeywords}
-functional programming, teaching
-\end{IEEEkeywords}
+\category{K.3.2}{Computing Milieux}{Computers and Education}[Computer and Information Science Education]
+\category{D.1.1}{Software}{Programming Techniques}[Applicative (Functional) Programming]
+
+\terms{Languages}
+
+\keywords{functional programming, teaching}
 
@@
 programming and information storage models. As a result, during the last few
-years, the practical software engineering field is witnessing a noticeable shift
-towards functional programming. Scripting languages, notably Javascript and
-Ruby, pioneered the introduction of functional concepts, such as closures and
-lambda functions, to mainstream programming. A new wave of programming
-languages, developed to overcome the expressiveness and complexity limitations
-exhibited in mainstream languages, have promoted functional constructs, such as
-type safe pattern matching, higher order functions and single assignment
-variables, to first class citizens (Scala, {\sc c\#}).  New, functional
-languages have emerged to fill in the remaining gaps (F\#, Clojure, Racket),
-often introducing significant advancements in their field of specialisation
-(such as Erlang in distributed fault-tolerant systems).  Finally, large scale
-information processing systems such as Map/Reduce~\cite{Dean04} and domain
-specific languages such as {\sc linq} have integrated functional concepts to
-ease the expression of masively parallel computations.
+years, the leading edge of practical software engineering field is witnessing a
+slight but noticeable shift towards functional programming ({\sc
+fp}).\footnote{See, for example, the reactive manifesto:
+\href{http://www.reactivemanifesto.org/}{http://www.reactivemanifesto.org/}}
+Scripting languages, notably Javascript and Ruby, pioneered the introduction of
+functional concepts, such as closures and lambda functions, to mainstream
+programming. A new wave of programming languages, developed to overcome the
+expressiveness and complexity limitations exhibited in mainstream languages,
+have promoted functional constructs, such as type safe pattern matching, higher
+order functions and single assignment variables, to first class citizens (Scala,
+{\sc c\#}).  New, functional languages have emerged to fill in the remaining
+gaps (F\#, Clojure, Racket), often introducing significant advancements in their
+field of specialisation (such as Erlang in distributed fault-tolerant systems).
+Finally, large scale information processing systems such as
+Map/Reduce~\cite{Dean04} and reactive programming~\cite{Meije12} and domain
+specific languages such as {\sc linq}~\cite{Meije11} have integrated functional
+concepts to ease the expression of massively parallel computations.
 
-Broadly speaking, functional programming is a style of programming in which the
-primary method of computation is the application of functions to arguments.
-Among other features, functional languages offer a compact notation for writing
-programs, powerful abstraction methods for structuring them, and a simple
-mathematical basis that supports reasoning. Many of the advanced techniques in
-modern functional languages, such as monads and catamorphisms, are closely based
-on principles from category theory such as functors, initial algebras, monads
-and Kleisli categories.
+Strictly speaking, {\sc fp} is a style of programming in which the primary
+method of computation is the application of functions to arguments.  Among other
+features, functional languages offer a compact notation for writing programs,
+powerful abstraction methods for structuring them, and a simple mathematical
+basis that supports reasoning. Many of the advanced techniques in modern
+functional languages, such as monads~\cite{Wadle93} and
+catamorphisms~\cite{Meije91}, are closely based on principles from category
+theory such as functors, initial algebras, monads and Kleisli categories.
 
-While functional programming has been taught for long in computer science
-departments~\cite{Joost93}, curricula tend to emphasize functional programming
-theory rather than practical applications. Special programming languages are
-used to teach functional programming specific concepts, while little connection
-is made to how those concepts can be transfered to non-functional languages.
-The course at TU Delft attempted to teach functional programming principles in the
-classroom and then expect students to apply those concepts in the (non-purely
-functional) language of their choice. The course had a very strong teaching by
-example focus: students were expected to participate in both in-classroom
-exercises, homework assignments and implement a real world system as a final
-project.
+While {\sc fp} has been taught for long in computer science
+departments~\cite{Joost93}, curricula tend to emphasize {\sc fp} theory rather than practical applications. Special programming languages are
+used to teach {\sc fp} specific concepts, while little connection
+is made to how those concepts can be transfered to solving practical software
+engineering problems.
+
+The presented course runs at TU Delft for two consecutive years. Our goal, and
+the course's moto, is to teach students to  ``think like ({\sc fp})
+fundamentalists and code like hackers''. The method we follow is teaching
+{\sc fp} principles in the classroom and then expecting students
+to apply those concepts in challenging practical projects. The course has a very
+strong teaching by example focus: students are expected to participate in both
+in-classroom exercises, homework assignments and implement a real world system
+as a final project, all within a half-semester (6 -- 8 weeks).
 
@@
 
-The functional programming course is elective at the MSc level. Students
-from all departments in the Electrical Engineering, Mathematics and Computer
-science faculty are allowed to participate. For practical reasons, the
-enrollment to the course was limited to 15 participants. This permitted
-us to get to know the students individually and to provide them with
-personalized support.
+The {\sc fp} course is elective at the MSc level. Students from
+all departments in the Electrical Engineering, Mathematics and Computer science
+faculty are allowed to participate. In the first year, we restricted the
+enrollment to the course to 15 participants; we lifted this
+restriction during the second year, as we found that the format could allow for
+more students to participate.
 
-From the students that enrolled, most have received formal introduction to
+From the students that enroll, most have received formal introduction to
 imperative and object oriented programming in their bachelors curriculum, while
-through participation to other courses they had limited exposure to functional
-programming (i.e. Map/Reduce in the web information systems
-course). Two of the students were majoring in computer engineering, which meant
-that their programming experience was restricted. Several students also had work
-experience as programmers as part of their industrial placement or through
-participation to start up ventures.
+through participation to other courses they had limited exposure to {\sc fp}
+(i.e. Map/Reduce in the Web Information Systems course). Some students are
+majoring in computer engineering, which meant that their programming experience
+was more restricted. Most students also have work experience as programmers as
+part of their industrial placement or through participation to start up
+ventures.
 
-The diverse backgrounds of the students guaranteed that no generic
+The diverse backgrounds of the students guarantees that no ge\-ne\-ric
 introduction to the topic would be sufficient. For this reason, we decided
 to adopt a hands-on-first approach; the students would have to learn by
-flexing their programming muscles, instead of being introduced to 
-the theory through toy examples.
+flexing their programming muscles, instead of being gently introduced to 
+the theory through toy examples. During the lectures, we made clear to the
+students that it is the delta in learning that we expect them to
+demonstrate rather than impressive technical achievements.
 
@@
 
-Functional programming is arguably one of the oldest programming paradigms. In it is
-purest form, it is based on a minimalistic theoretical background
-($\lambda$-calculus). Relatively recent
-formulations~\cite{Meije91, Wadle93} also introduced concepts from category
-theory. While the every day use of functional programming does not necessarily
-require the programmer to be aware of the theory, understanding it usually leads
-to more elegant algorithmic solutions. Functional programming teaching is also
-associated with advanced type systems; indeed, the flagship functional
-programming languages (Haskell and the {\sc ml} family) both offer very
-sophisticated support for types. Consequently, teaching the full spectrum of
-functional programming theory and techniques would have been impossible in a
-half-semester course; instead we decided to focus on practical aspects of data
-processing, transformations and state representation using functional
-techniques.
+{\sc fp} is arguably one of the oldest programming paradigms. In
+it is purest form, it is based on a minimalistic theoretical background
+($\lambda$-calculus). Relatively recent formulations~\cite{Meije91, Wadle93}
+also introduced concepts from category theory. While the every day use of
+{\sc fp} does not necessarily require the programmer to be aware
+of the theory, understanding it usually leads to more elegant algorithmic
+solutions. Teaching {\sc fp} is also associated with advanced type
+systems; indeed, the flagship {\sc fp} languages (Haskell, Scala) both offer very sophisticated support for types. Consequently,
+teaching the full spectrum of {\sc fp} theory and techniques is impossible in a half-semester course; instead we focus on
+practical aspects of data processing, transformations and state representation
+using functional techniques.
 
-\subsection{Choosing the appropriate programming language}
+\subsection{Choosing a programming lan\-guage}
 
@@
 application, the less important students feel the taught concepts are.
-Fortunately, most functional programming concepts can be expressed cleanly by
-several widespread languages: for example, Javascript, Ruby and {\sc c\#} have
+Fortunately, most {\sc fp} concepts can be expressed cleanly by
+several widespread languages: for example, Ja\-va\-script, Ruby and {\sc c\#} have
 closures, first class and higher order functions, while a number of emerging
@@
 with the existing literature, there is no practical reason that necessitates
-teaching strictly in a language like Haskell or {\sc ml}.
+teaching in a strict {\sc fp} language like Haskell or {\sc ml}.
 
 The above led us to not choose any particular language for the course.  During
-the lectures, the demonstration languages where Haskell and {\sc c\#}, while at
-the labs we used Scala. We actively encouraged students to use the language of
+the lectures, the demonstration languages where Haskell, {\sc c\#} and {\sc php}, while at the labs we used Scala. We actively encouraged students to use the language of
 their choice on the platform of their choice to carry out homework assignments
@@
 
-The high level goal of the course was to teach the principles of functional
-programming, and the corresponding Category theoretical principles. 
-More specifically, the educational purposes of the course were:
+The high level goal of the course was to teach the principles of {\sc fp}, and the corresponding Category theoretical principles. More specifically, the educational purposes of the course were:
 
@@
 
-  \item To introduce students to basic functional programming concepts,
+  \item To introduce students to basic {\sc fp} concepts,
     higher order functions, monads and advanced type systems.
@@
 
-The course consisted of a series of lectures, of which two where invited
-lectures by an external instructor, and laboratory sessions. In total, the
-course consisted of 14 hours of lectures and 8 hours of lab work.
+The course consists of a series of lectures. In total, the
+course consisted of 15 hours of lectures. 
 The students
-also had to carry out a series of homework assignments and a final project.
-The evaluation would be based on student performance on the final assignment. 
-The lectures took place in a quarter (half semester), during 3 intense 
-weeks; the students had another 6 weeks to finish their final projects.
+also have to carry out a series of optional homework assignments and a final project.
+The evaluation is based on student performance on the final assignment. 
+The lectures take place in a quarter (half semester), during 6 intense 
+weeks; the students have another 2-3 weeks to finish their final projects.
 
@@
 
-  \item Real time graph visualisation on steaming data. The
-    particular example that students worked on was visualizing community
-    structures for Github projects.
+  \item Real time graph visualisation on steaming data.
 
-  \item Multi-source real time data processing. Students worked on a 
-    programming language popularity index, based on data from the Github
-    event stream and the StackOverflow tag cloud {\sc api}.
+  \item Haskell constructs (e.g. the Prelude or abstract data types) in Javascript or even {\sc php}.
 
-  \item Implementation of Haskell constructs in Javascript. Students implemented
-    the Haskell Prelude (basic functions for list manipulation).
+  \item Interpreters and byte code converters between languages.
 
-  \item Implementation of Map/Reduce algorithms in Haskell. Students installed
-    and configured cloud Haskell and implemented simple Map/Reduce based
-    algorithms in a distributed setting.
+  \item Simple constraint solvers and machine learning algorithms.
 
-  \item Implementation of simple constraint solvers.
-
-  \item Machine learning algorithms. Students implemented Na\"ive Bayes 
-    classifiers and K-means clustering algorithms.
+  \item Asynchronous request processing in network servers
 
@@
 
-To counterbalance the differences in student experience level, 
-the assignments were open ended; students could drive their projects as
-far as they where willing to.
-The deliverables for the assignments were a repository
-with the source code and short report describing the solution that the
-students came up with. The students were also required to present a working
-demo by the end of the course time.
+To counterbalance the differences in student experience level, the assignments
+are open ended; students can drive their projects as far as they are willing to,
+as long as the experience is pleasant enough to allow them to learn. The deliverables for the
+assignments are a repository with the source code and short report describing
+the solution that the students came up with. The students are also required to
+present a working demo by the end of the course time.
 
@@
 
-The purpose of the lab sessions where to help students carry out their final
-assignments and further explain concepts that might not be clear to students. 
-Students convened every week, presented their progress and discussed with 
-each other and with the lab demonstrator design and practical issues that
-arised in the implementation of their projects. To further motivate students
-to work on their projects, one of the lab sessions was converted to a full
-day coding sprint.
+In the first year it was offered, the course included lab sessions.  The purpose
+of the lab sessions were to help students carry out their final assignments and
+further explain concepts that might not be clear to them.  Students convened
+every week, presented their progress and discussed with each other and with the
+lab demonstrator design and practical issues that hindered the implementation of
+their projects. To further motivate students to work on their projects, one of
+the lab sessions was converted to a full day coding sprint. However, we found
+out that participation was diminishing and the sessions were not of much interest to the students,
+who preferred to have more time to work on the projects. We therefore canceled
+the lab sessions in the course's second year.
+
+\subsection{Organizing the work}
+
+The higher number of students in the second year of the project made the
+organization of student work challenging. To better organize and observe the
+student's work, we made active use of Github. Github is a code hosting site that
+includes social networking features (such as notifications on events, following
+of projects etc) and a dashboard to observe them. We created an organization for
+the course\footnote{\href{https://github.com/organizations/fptudelft}{https://github.com/organizations/fptudelft}} and in it we
+created repositories for the projects shared among all students registered on
+the course. We also created a specialized repository for answering general
+questions about the course, using the repository's issue tracking facility.
+
+Github's notification functionality enabled us and the teaching assistants to
+answer questions about the student projects in almost real time. We could also
+observe work done by the students while it was done and this allowed us to
+formulate an early opinion on their progress.
+We then informally discussed this with the students during idle time (e.g.
+lecture breaks), giving two types of feedback: praise to students who were doing good, help and support to students that were struggling.
 
@@
 
-The focus of teaching was to present functional programming from a practical
+The focus of teaching was to present {\sc fp} from a practical
 aspect; we were mostly interested to teach students what they can do rather than
 the theory behind functional constructs. An important distinction that was made
-early on was that the world is imperative; therefore while functional
-programming can be a great tool for thinking about a problem, a pure functional
+early on was that the world is imperative; therefore while {\sc fp} 
+ can be a great tool for thinking about a problem, a pure functional
 solutions cannot reflect on the real world. As a consequence, it is usually best
 to mix programming paradigms, i.e. imperative or object-oriented programming for
-state representation and functional programming for data processing. In our
+state representation and {\sc fp} for data processing. In our
 experience, this distinction helped students understand that they are already
-doing functional programming without noticing; enforcing data immutability to
+doing {\sc fp} without noticing; enforcing data immutability to
 handle multi-threading problems or passing function arguments in scripting
-languages are applications of functional programming principles. The moment
-students realized this fact, their attention was captured to the lectures.
+languages are applications of {\sc fp} principles. The moment
+students realized this, their attention was captured to the lectures.
 
 During both lectures and labs, all non-whiteboard examples where actually
-entered in a real programming environment (LinqPad or the Scala command line
-read-evaluate loop) and the results of the evaluation where discussed with the
-students. Participation was encouraged by modifying the examples and asking the
-students to perform the evaluation, before the example was executed. The
-examples varied from list manipulation with higher-order functions to 
-a toy actor model example.
+entered in a real programming environment (LinqPad in {\sc c\#} or the Scala
+command line read-evaluate loop) and the results of the evaluation where
+discussed with the students. Participation was encouraged by modifying the
+examples and asking the students to perform the evaluation, before the example
+was executed.
+The examples varied from list manipulation with higher-order
+functions to a toy actor model example.
+Moreover, we made extensive use of the whiteboard to demonstrate abstract
+concepts like type inference and combinators. Typically, we would
+write a function signature in Haskell and then expected students to come up 
+with its implementation and vice versa.
 
-The proverbial ``six ways to understand monads'' included presentations of the
-formal theoretical concepts, visualizations of monads as blenders (based on the
-monadic property that once a value is captured, it is then contained),
-applications in Haskell for encapsulating state (through the \textsf{Maybe}
-monad), applications to generic programming in {\sc c\#}, applications to {\sc
-linq} and the reactive extensions frameworks, and, finally, a whiteboard
-exercise where students were prompted to participate in the ad-hoc specification
-of Haskell's \textsf{IO} monad from basic principles. Once again, the examples
-enabled the students to understand the composition and state containment
-properties of an otherwise difficult to comprehend theoretical concept.
+Of particular interest to the students was the presentation of monads. The
+students were teased that they will learn monads in no less than 6 different
+ways. Those included presentations of the formal theoretical concepts,
+visualizations of monads as blenders (based on the monadic property that once a
+value is captured, it is then contained), applications in Haskell for
+encapsulating state (through the \textsf{Maybe} monad), applications to generic
+programming in {\sc c\#}, applications to {\sc linq} and the reactive extensions
+frameworks, and, finally, a whiteboard exercise where students were prompted to
+participate in the ad-hoc specification of Haskell's \textsf{IO} monad from
+basic principles. Once again, the examples enabled the students to understand
+the composition and state containment properties of an otherwise difficult to
+comprehend theoretical concept.
 
@@
 \label{fig:teapots}
-\subfigure[Input drawing, rendered with Java graphics]{
-\includegraphics[scale=0.173]{teapot-reference.png}
+\subfigure[Input model]{
+\includegraphics[scale=0.14]{reference.png}
 \label{fig:teapot-reference}
 }
-\subfigure[Decomposition in Scala, rendered with Java graphics]{
-\includegraphics[scale=0.17]{teapot-java-canvas.png}
+\subfigure[Decomposed model]{
+\includegraphics[scale=0.20]{decomposed.png}
 \label{fig:teapot-scala}
 }
-\subfigure[Decomposition in Scheme, rendered with appropriately positioned {\sc html} \textsf{div}
-elements]{
-\includegraphics[scale=0.195]{teapot-html-divs.png}
-\label{fig:teapot-html-divs}
+\subfigure[Detail: The teapot's handle]{
+\includegraphics[scale=0.20]{detail.png}
+\label{fig:teapot-scala}
 }
-\subfigure[Decomposition in Javascript, rendered with {\sc html5} graphics]{
-\includegraphics[scale=0.295]{teapot-html-canvas.png}
-\label{fig:teapot-html-canvas}
-}
-\caption[]{Example results from the teapot exercise.}
+
+\caption[]{The teapot input and decomposition results}
 
@@
 
-  \item Most modern programming languages can express functional programming
-    constructs.
+  \item Most modern programming languages can express {\sc fp} constructs.
 
-  \item Functional programming works best in data transformation scenarios.
+  \item {\sc fp} works best in data transformation scenarios.
 
@@
 
-The exercise consisted of rendering the Utah Teapot~\cite{Torre06} using any
-graphics primitive of the student's choice using only right triangles (with one
-horizontal leg).  In its
-core, the exercise required students to decompose a collection of arbitrary triangles, which
-comprised the input Utah Teapot model, to a collection of right triangles. The students
-had to come up with the decomposition method (using some simple analytic
-geometry, this was not a course on 3D graphics), a decomposition termination criterion to stop the decomposition when
-triangles were too small to be rendered on screen and a method to recursively
-apply the above mentioned transformations on the input data. As always, the
-students could decide the implementation language of their choice. The exercise
-was given as a mid-week homework assignment, between the Monday and the Friday
-lectures.
+The first time the course was given, the exercise consisted of rendering the
+Utah Teapot~\cite{Torre06} using any graphics primitive of the student's choice
+using only right triangles with one horizontal leg (see
+Figure~\ref{fig:teapots}). In its core, the exercise required students to
+decompose a collection of arbitrary triangles, which comprised the input Utah
+Teapot model, to a collection of right triangles~\cite{Beckm13}. The students
+had to come up with the decomposition method (using analytical geometry), a decomposition termination criterion to
+stop the decomposition when triangles were too small to be rendered on screen
+and a method to recursively apply the above mentioned transformations on the
+input data. As always, the students could decide the implementation language of
+their choice. The exercise was given as a mid-week homework assignment, between
+the Monday and the Friday lectures.
 
-The student's response was overwhelming. Even though it was made clear that the
-exercise would not contribute to the final grade, the will to apply the data
-processing techniques taught during the lectures motivated the students to work
-very hard. The fact that they were instructed to use their favourite language
-enabled them to focus on decomposing the problem in a series of testable
-calculation steps rather than meddling with the intricacies of learning a new 
-language. In spite of the advice, students were eager to test-drive the newly
-acquired knowledge using a functional language; apart from {\sc c\#} and
-Javascript, solutions were also provided in Scala, Scheme and Haskell. Example
-renderings produced by student programs can be seen in Figure~\ref{fig:teapots}.
+During the second run of the course, the exercise was optimizing decomposition
+rather than implementing it.
+We provided the students with a basic implementation in {\sc php} and asked them to optimize the code by applying new algorithmic techniques and taking into account new language features implemented in a
+pre-release version of Facebook's {\sc php} language version (codenamed Hack).
+To further motivate them, we converted the exercise to a programming
+contest with prizes for the winning team.
+
+Both times, the students' response was overwhelming. Even though it was made
+clear that the exercise would not contribute to the final grade, the will to
+apply the data processing techniques taught during the lectures motivated the
+students to work very hard. In the first year, the fact that they were
+instructed to use their favourite language enabled them to focus on decomposing
+the problem in a series of testable calculation steps rather than meddling with
+the intricacies of learning a new language. In spite of the advice, students
+were eager to test-drive the newly acquired knowledge using a functional
+language; apart from {\sc c\#} and Javascript, solutions were also provided in
+Scala, Scheme and Haskell. In the second year, the students learned the
+value of clean design and breaking code changes to obtain massive
+optimizations~\cite{Gousi13}.
 
@@
 According to many students, the Map/Reduce data analysis framework was what
-drove them to attend the course. The lecture on Map/Reduce featured an invited
+drove them to attend the course. In the first year, the lecture on Map/Reduce featured an invited
 talk by Prof. Ralf L\"ammel, who presented his formulation of Map/Reduce using
@@
 dedication. The source code to all student projects and the teapot
-exercise can be found at (or linked from) \url{http://swerl.tudelft.nl/bin/view/Main/FunctionalProgrammingCourse}. This work is partially supported by
-Marie Curie {\sc ief} 298930 -- {\sc sefunc}.
+exercise can be found at (or linked from) \href{https://github.com/organizations/fptudelft}{https://\-github.com\-/organizations\-/fptudelft}.
+This work is partially supported by Marie Curie {\sc ief} 298930 -- {\sc sefunc}.
 
 \bibliography{paper}
-\bibliographystyle{ieeetr}
+\bibliographystyle{abbrv}
 \end{document}
